draft

### 2025-05-20 draft.tex
fix minor typos
docs: Update draft text

% Created 2025-05-20 Tue 10:41
% Intended LaTeX compiler: pdflatex
\documentclass{article}

@@

\begin{figure}[htbp]
\centering
\includesvg[width=.9\linewidth]{./figs/cpmeasure_overview}
\caption{\label{fig:overview}cp\_measure generates features from images by using the local information in every region of interest. It is commonly used to featurise the pairwise combination of all the available channels and objects. The resultant matrices represent the entire experiment and can be studied using machine and deep learning methods.}
\end{figure}

@@

We first performed the numerical validation of cp\_measure, relative to the original CellProfiler features. For this we collected 300 images corresponding to 150 perturbations from the JUMP dataset, selecting the most significant phenotypes for a given measurement each. To ensure that we are using identical object masks, we segmented these images to obtain the cells and nuclei using CellProfiler, providing object masks and their associated features. Next, we applied cp\_measure on these masks with the original images and mapped the features from cp\_measure to CellProfiler. Lastly, we calculated a linear fit for the matched features and calculated their \(R^2\) value, indicating how well it fits a linear slope.

The validation of our cp\_measure features is shown on \ref{fig:cp_vs_cpmeasure}. The left panel shows examples of the comparison of our features against CellProfiler's. The straight lines demonstrate the recapitulation of measurements from our implementation. A few data points fall outside the diagonals, which hint at some edge-cases are treated differently by either tool. The panel on the right shows the \(R^2\) value of a linear interpolation. Given that this value is directly correlated to the correctness of the implementation, we can see that most of our measurements have a linear relationship, regardless of whether the masks were for nuclei or cytosols. This result provides reassurance that cp\_measure can be confidently in cases where CellProfiler would be used.
\subsection{Results}
\label{sec:orge5b5c6b}
We showcase a couple of use-cases in which cp\_measure makes our machine-learning workflows faster and integrate better with existing tools.
\subsubsection{Astrocytes 3D data}
\label{sec:org447090b}

@@

Figure \ref{fig:astrocytes} shows an example image and object masks alongside the Shapley values of a classifier trained on cp\_measure features. Our results showcased the axis length of the cell to be a major indicator of phenotypic effect, implying that cells became more elongated on their minor axis. While it was possible to CellProfiler for this task, cp\_measure made it much faster and simpler to implement, specially since the segmentation masks were already provided, which is quite common in public datasets.
\subsubsection{Spatial transcriptomics}
\label{sec:org5711d86}
A key advantage of providing these measurements as a standalone Python package is their ease of integration into diverse analytical workflows, which otherwise would require substantial adaptation to the standard CellProfiler environment. The recent proliferation of black-box foundation models trained solely on morphological data highlights morphology as a highly informative and predictive modality. However, the feature vectors produced by these models are typically not interpretable, preventing direct biological assessment. In contrast, classical morphological measurements yield explicit, interpretable readouts -- for instance, the co-localization of fluorescent markers -- facilitating clear biological interpretations.

>>>>

% Created 2025-05-20 Tue 11:00
% Intended LaTeX compiler: pdflatex
\documentclass{article}

@@

\begin{figure}[htbp]
\centering
\includesvg[width=.9\linewidth]{./figs/cpmeasure_overview}
\caption{\label{fig:overview}cp\_measure generates features from images by using the local information in every region of interest. It is commonly used to featurize the pairwise combination of all the available channels and objects. The resultant matrices represent the entire experiment and can be studied using machine and deep learning methods.}
\end{figure}

@@

We first performed the numerical validation of cp\_measure relative to the original CellProfiler features. For this we collected 300 images corresponding to 150 perturbations from the JUMP dataset, selecting the most significant phenotypes for a given measurement each. To ensure that we are using identical object masks, we segmented these images to obtain the cells and nuclei using CellProfiler, providing object masks and their associated features. Next, we applied cp\_measure on these masks with the original images and mapped the features from cp\_measure to CellProfiler. Lastly, we calculated a linear fit for the matched features and calculated their \(R^2\) value, indicating how well it fits a linear slope.

The validation of our cp\_measure features is shown on \ref{fig:cp_vs_cpmeasure}. The left panel shows examples of the comparison of our features against CellProfiler's. The straight lines demonstrate the recapitulation of measurements from our implementation. A few data points fall outside the diagonals, which hint at some edge-cases are treated differently by either tool. The panel on the right shows the \(R^2\) value of a linear interpolation. Given that this value is directly correlated to the correctness of the implementation, we can see that most of our measurements have a linear relationship, regardless of whether the masks were for nuclei or cytosols. This result provides evidence that cp\_measure can be confidently in cases where CellProfiler would be used, and the insights are unlikely to change.
\subsection{Results}
\label{sec:orge5b5c6b}
\subsubsection{Astrocytes 3D data}
\label{sec:org447090b}

@@

Figure \ref{fig:astrocytes} shows an example image and object masks alongside the Shapley values of a classifier trained on our features. This indicates that the major axis length of the cell to is an indicator of phenotypic effect over the course of the experiment, implying that cells became more elongated on their minor axis. Though unsurprising, as we expect astrocytes to extend over time, cp\_measure was able to recover this with only a few lines of code.
\subsubsection{Spatial transcriptomics}
\label{sec:org5711d86}
A key advantage of providing these measurements as a standalone Python package is their ease of integration into diverse analytical workflows, which otherwise would require substantial adaptation to the standard CellProfiler environment. The recent proliferation of black-box foundation models trained solely on morphological data highlights morphology as a highly informative and predictive modality. However, the feature vectors produced by these models are typically not interpretable, preventing direct biological assessment. In contrast, classical morphological measurements yield explicit, interpretable readouts -- for instance, the co-localization of fluorescent markers -- facilitating clear biological interpretations.
